The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
done with 3

@@

\item Each of the values {{-a + b}} and {{(1/a) * b}} can be viewed as a
  function that asks for an {{a}} and then produces a {{b}}. When viewed as
  functions, we write these types as {{a -o+ b}} and {{a -o* b}}
  respectively. Alternatively we can view these values as first producing a
  {{b}} and then demanding an {{a}} and in that perspective they correspond
  to delimited continuations. Evidently, as the discussion above suggests,
  these two notions of functions are not the same at all and should not be
  conflated. Sec.~\ref{sub:hof} discusses this point in detail.

@@

@@

The interesting thing about the semantics is that it represents a reversible
abstract machine. In other words, we can compute the start state from the
stop state by changing the reductions {{|-->}} to run backwards
{{<--|}}. When running backwards, we use the isomorphism represented by a
combinator {{c}} in the reverse direction which we indicate by writing
{{c{dagger}}}. For example, {{distrib{dagger} = factor}} and so on.

@@

For a monoidal category to be compact closed, the maps {{eta}} and {{eps}}
must satisfy a coherence condition that is usually visualized as follows
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
In the case of negatives, the condition amounts to checking that reversing
direction twice is a no-op. In the case of fractionals, the condition amounts
to checking that creating values {{alpha}} and {{1/alpha}} and immediately
unifying them is also a no-op. Both checks are straightforward.

@@

We have omitted some the commutativity and associativity shuffling to
communicate the main idea. We are given a value of type {{a}} which we embed
into {{a+0}} and then {{a + (c - c)}}. This can be reassociated into 
{{(a+c) - c}}. The component {{a+c}} which until now is just an appropriately 
tagged value of type {{a}} can be given to {{f}} which transforms it to
something of type {{b+c}}. If the result is in the {{b}}-summand, it is produced
as the answer; otherwise the result is in the {{c}}-summand; {{eps+}} is used to
make it flow backwards to be fed to the {{eta+}} located at the beginning of the 
sequence. Iteration continues until a {{b}} is produced.

@@

{{b <-> -(-b)}}

\begin{center}
  \includegraphics{diagrams/double_neg.pdf}
\end{center}

@@

Mention LL and {{a -o b}}.

\begin{center}
  \includegraphics{diagrams/function.pdf}
\end{center}

Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
Sec. \ref{sec:hof} for a discussion of the additive functions
vs. multiplicative functions.

@@

\scalebox{0.8}{
    \includegraphics{diagrams/compose.pdf}
}
\end{multicols}
This is also equivalent to sequencing both the computation blocks.

@@

\begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/apply1.pdf}
}
\end{center}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/apply2.pdf}
}
\end{center}
\end{multicols}

%%%%%
\paragraph*{Delimited continuation}

\begin{center}
  \includegraphics{diagrams/delimc.pdf}
\end{center}

In a conventional language, higher-order functions capture scope and can be
represented as first-class values. In {{langRevEE}}, we have two notions of
functions both of which share similarities with conventional higher-order
functions.

We write these type
written {{a-o*a}} whose witness may be thought of as the identity function at
the type {{a}}.  For example, consider a datatype {{color = R|Gr|B}}, and let
us consider the following manipulations:

\begin{itemize}
\item Using the fact that {{1}} is the multiplicative unit, generate
  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
  the components of the above tuple. The resulting value is
  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
  {{alpha2}} are fresh logic variables;
\item Using the fact that {{*}} is associative and commutative, we can
  rearrange the above tuple to produce the value:

@@

Dually, the additive reading of the above manipulations correspond to
functions of the form {{a-o+a}}, witnessing isomorphisms of the form
{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
between two subcomputations, \emph{only one of which exists} at any point,
i.e., they capture the essence of coroutines. Additive and multiplicative
functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
that in a universe in which information is not guaranteed to be preserved by
the computational infrastructure, the above slicing and dicing of functions
would make no sense. Once the computational framework guarantees that
information is maintained, it becomes possible to break dependencies,
generate debts and constraints, and let these flow as first-class values,
because they \emph{must} eventually be satisfied.

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions specific to Negatives and Fractions}
\label{sec:specific-constructions}

\emph{Negation distributes over {{+}}. }

{{-(b1 + b2) <-> (-b1) + (-b2)}}

\begin{center}
  \includegraphics{diagrams/dist_neg_plus.pdf}
\end{center}

\emph{Lifting an operation of positive types to negated types}

\begin{multicols}{2}
Given {{c : b1 <-> b2}}

@@

Annihilation and non-termination describe undefined program
execution. While non-termination characterizes an iteration that fails
to terminate, annihilation characterizes a constraint that has no
fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
annihilating only the program states that do not match the boolean
satisfiability constraint.

\emph{Annihilation.}  
With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
which gives us the ability to find the fixpoint of a constraint.  If
the constraint has no satisfying values, there is no fixpoint possible
-- we use the term \emph{annihilation} to describe the corresponding
undefined state of the program.

While the constructions in Sec \ref{sec:neg} apply to both the
{{(0,+)}} and {{(1,*)}} monoids, each construction presented here
applies to a specific monoid.

\begin{enumerate}

\item 
\emph{Lifting negation out of {{*}}. }

@@

\item
\emph{Multiplying Negatives}

@@

This isomorphism can be built using \emph{involution} constructed in
Sec~\ref{sec:neg} and the isomorphisms above, and correspond to the
algebraic manipulation:

@@

\item 
\emph{Multiplying Fractions}

@@

\item 
\emph{Adding the numerator}

@@

\item 
\emph{Fraction Addition}

@@

@@

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit {{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}, denoted {{f^{dagger}}}. At this point we have,
the top half of the construction below:

>>>>

\newcommand{\fname}[1]{\ulcorner #1 \urcorner}
\newcommand{\fconame}[1]{\llcorner #1 \lrcorner}

@@

\item Each of the values {{-a + b}} and {{(1/a) * b}} can be viewed as a
  function that asks for an {{a}} and then produces a {{b}}. When viewed as
  functions, we write these types as {{a -o+ b}} and {{a -o* b}}
  respectively. Alternatively we can view these values as first producing a
  value of type {{b}} and then demanding an {{a}} and in that perspective
  they correspond to delimited continuations. Evidently, as the discussion
  above suggests, these two notions of functions are not the same at all and
  should not be conflated. Sec.~\ref{sub:hof} discusses this point in detail.

@@

\paragraph*{Adjoint.} 
An important property of the language is that every combinator {{c}} has an
adjoint {{c{dagger}}} that reverses the action of {{c}}. This is evident by
construction for the primitive isomorphisms. For the closure combinators, the
adjoint is homomorphic except for the case of sequencing in which the order
is reversed, i.e., {{(c1 (;) c2){dagger} = (c2{dagger}) (;) (c1{dagger}) }}.

@@

The interesting thing about the semantics is that it represents a reversible
abstract machine. In other words, we can compute the start state from the
stop state by changing the reductions {{|-->}} to run backwards
{{<--|}}. When running backwards, we use the isomorphism represented by a
combinator {{c}} in the reverse direction, i.e., we use the adjoint
{{c{dagger}}}.

@@

For a monoidal category to be compact closed, the maps~{{eta}} and~{{eps}}
must satisfy a coherence condition that is usually visualized as follows:
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}
In the case of negatives, the condition amounts to checking that reversing
direction twice is a no-op. In the case of fractionals, the condition amounts
to checking that creating values {{alpha}} and {{1/alpha}} and immediately
unifying them is also a no-op. Both checks are straightforward.

@@

\noindent We have omitted some the commutativity and associativity shuffling to
communicate the main idea. We are given a value of type {{a}} which we embed
into {{a+0}} and then {{a + (c - c)}}. This can be reassociated into 
{{(a+c) - c}}. The component {{a+c}} which until now is just an appropriately 
tagged value of type {{a}} can be given to~{{f}} which transforms it to
something of type {{b+c}}. If the result is in the {{b}}-summand, it is produced
as the answer; otherwise the result is in the {{c}}-summand; {{eps+}} is used to
make it flow backwards to be fed to the {{eta+}} located at the beginning of the 
sequence. Iteration continues until a {{b}} is produced.

@@

In a symmetric compact closed category, we can build isomorphisms that the
dual operation is an involution. Specifically, we get the isomorphisms 
{{b <-> -(-b)}} and {{b <-> (1/(1/b))}}. For the additive case, the 
isomorphism is defined as follows:

{{ (id (+) eta+) (;) (swap+ (+) id) (;) (id (+) eps+) }}

\noindent where we have omitted the 0 introduction and elimination. The idea
is as follows: we start with a value of type {{b}}, embed it into {{b+0}} and
use {{eta}} to create something of type {{b + (-(-b) + (-b))}}. This is possible
because {{eta}} has the polymorphic type {{-a + a}} which can be instantiated
to {{-b}}. We then reshuffle the type to produce {{-(-b) + (-b + b)}} and cancel
the right hand side using {{eps+}}.  The construction for the multiplicative 
case is identical and omitted.

%% {{b <-> -(-b)}}
%% 
%% this is the wrong diagram: see lemma 4.17 in selinger's paper
%% \begin{center}
%%   \includegraphics{diagrams/double_neg.pdf}
%% \end{center}

@@

Although these constructions are also standard, they are less known and they
are particularly important in our context: we devote a little more time to
explain them. Our discussion is mostly based on Abramsky and Coecke's article
on categorical quantum mechanics~\cite{abramsky-2008}.

In a compact closed category, each morphism {{f : a <-> b }} can be given a
\emph{name} and a \emph{coname}. For the additive fragment, the name
$\fname{f}$ has type {{0 <-> (-a + b)}} and the coname $\fconame{f}$ has type
{{a + (-b) <-> 0}}. For the multiplicative fragment, the name $\fname{f}$ has
type {{1 <-> ((1/a) * b))}} and the coname $\fconame{f}$ has type 
{{(a * (1/b)) <-> 1}}. Intuitively, this means that for each morphism, 
it is possible to construct, from ``nothing,'' an object in the category that 
denotes this morphism, and dually it is possible to eliminate this object.
The construction of the name and coname of {{c : b1 <-> b2}} in the additive case 
can be visualized as follows:

\begin{center}
  \includegraphics{diagrams/function.pdf}
\end{center}
\begin{center}
  \includegraphics{diagrams/delimc.pdf}
\end{center}

Intuitively the name consists of viewing {{c}} as a function and the coname
consists of viewing {{c}} as a delimited continuation.

In addition to being able to represent morphisms, it is possible to express
function composition. For the additive case, the composition is depicted
below:

@@

\scalebox{0.8}{
    \includegraphics{diagrams/compose.pdf}
}
\end{multicols}
which is essentially equivalent to sequencing both the computation blocks as
shown below:

@@

Applying a function consists of making it flow backwards to satisfy the demand 
of the function:
\begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/apply1.pdf}
}
\end{center}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/apply2.pdf}
}
\end{center}
\end{multicols}

Having reviewed the representation of functions, we now discuss the
similarities and differences between the two notions of functions and their
relation to conventional (linear) functions which mix additive and
multiplicative components. For that purpose, we use a small example.
Consider a datatype {{color = R|Gr|B}}, and let us consider the following
manipulations:
\begin{itemize}
\item Using the fact that {{1}} is the multiplicative unit, generate
  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
  the components of the above tuple. The resulting value is
  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
  {{alpha2}} are fresh logic variables;
\item Using the fact that {{*}} is associative and commutative, we can
  rearrange the above tuple to produce the value:

@@

Dually, the additive reading of the above manipulations correspond to
functions of the form {{a-o+a}}, witnessing isomorphisms of the form
{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
between two subcomputations, \emph{only one of which exists} at any point,
i.e., they capture the essence of coroutines.

It should be evident that in a universe in which information is not
guaranteed to be preserved by the computational infrastructure, the above
slicing and dicing of functions would make no sense. But linearity is not
sufficient: one must also recognize that the additive and multiplicative
spaces are different.

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additional Constructions}
\label{sec:specific-constructions}

The additional constructions below (presented with minimal commentary)
confirm that conventional algebraic manipulations in the mathematical field
of rationals do indeed correspond to realizable type isomorphisms in our
setting. The constructions involving both negative and fractional types are
novel.

\paragraph*{Negation distributes over {{+}}. }
One certainly expects that:
{{-(b1 + b2) <-> (-b1) + (-b2)}}
which can be realized as follows:
\begin{center}
  \includegraphics{diagrams/dist_neg_plus.pdf}
\end{center}

\paragraph*{Lifting an operation of positive types to negated types.}
Any operation on types can be applied to the negative versions of these
types. The idea is simply to reverse the flow of values and use the
adjoint of the operation:
\begin{multicols}{2}
Given {{c : b1 <-> b2}}

@@

\paragraph*{Lifting negation out of {{*}}.}
The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
and {{b2}} are flowing opposite to each other then it doesn't matter which is
direction is forwards and which is backwards.

@@

\paragraph*{Multiplying Negatives.}

@@

This isomorphism is a consequence of the fact that $-$ is an involution: it
corresponds to the algebraic manipulation:

@@

\paragraph*{Multiplying and Adding Fractions.}
An isomorphism witnessing:

@@

is straightforward. More surprisingly, it is also possible to construct
isomorphisms witnessing:

@@

@@

\amr{to merge}
Annihilation and non-termination describe undefined program
execution. While non-termination characterizes an iteration that fails
to terminate, annihilation characterizes a constraint that has no
fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
annihilating only the program states that do not match the boolean
satisfiability constraint.
\emph{Annihilation.}  
With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
which gives us the ability to find the fixpoint of a constraint.  If
the constraint has no satisfying values, there is no fixpoint possible
-- we use the term \emph{annihilation} to describe the corresponding
undefined state of the program.

@@

With a little work, we can modify this circuit to only annihilate values that
fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
detail, an instance of SAT is a function/circuit {{f}} that given some
boolean inputs returns {{true}} or {{false}} which we interpret as whether
the inputs satisfy the constraints imposed by the structure of {{f}}. Because
we are in a reversible world, our instance of SAT must be expressed as an
isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
below. Assuming that {{f}} is expressed as an isomorphism, we have enough
information to reconstruct the input from the output. This can be done by
using the adjoint of {{f}}. At this point we have, the top half of the
construction below:
